<span style="color:red;">Building a house with<br />Advanced Normalization Tools (*ANTs*)</span>

### 2014-10-07 ants2015build.Rmd
WIP: regular update
WIP: for cfn meeting

# Background

## Founding developers

<div align="center"><img width="1800" src="./figures/bant2.png" frameborder="0"></img></div>

BBA & NT

## Long-term collaborators

![](./figures/antscollab.jpg)

$+$ [neurodebian](http://neuro.debian.net/pkgs/ants.html), [slicer](http://www.slicer.org/), [brainsfit](https://github.com/BRAINSia/BRAINSTools), [nipype](http://nipy.sourceforge.net/nipype/), [itk](http://www.itk.org) and more ...

![](./figures/lion.png)

a pride: common way of doing things

... in a competitive world ...

## Definitions

* Registration $=$ estimate an "optimal" geometric mapping between image pairs or image sets (e.g. Affine)

* <span style="color:grey;"> Similarity $=$ a function relating one image to another, given a transformation (e.g. mutual information)</span>

* Diffeomorphisms $=$ differentiable map with differentiable inverse (e.g. "silly putty", viscous fluid)

* <span style="color:grey;">Segmentation $=$ labeling tissue or anatomy in images, usually automated (e.g. K-means)</span>

* Multivariate $=$ using many voxels or measurements at once (e.g. PCA, $p >> n$ ridge regression)</span>

* <span style="color:grey;">Multiple modality $=$ using many modalities at once (e.g. DTI and T1 and BOLD)</span>

* MALF: multi-atlas label fusion - using anatomical dictionaries to label new data

* <span style="color:grey;">Solutions to challenging statistical image processing problems usually need elements from each of the above</span>

## Image mapping & perception: 1878

![](./figures/galton.png)

* Francis Galton: Can we see criminality in the face?

* (maybe he should have used ANTs?)

## Initial scope

![](./figures/ants_initial_scope.png)

... just do a better registration (tell story) ...

## ANTs Lineage

![](./figures/lineage.jpg)

References: @Horn1981, @Gee1993, @Grenander1993, @Thompson2001, @Miller2002, @Shen2002, @Arnold2014, @Thirion1998, @Rueckert1999, @Fischl2012, @Ashburner2012

## Diffeomorphisms

<div align="center"><img width="1433" src="./figures/sillyputty.png" frameborder="0"></img></div>

plausible physical modeling of large, invertible deformations

"differentiable map with differentiable inverse"

## Fine-grained and flexible maps
![](./figures/highresdiffeos.jpg)

... to correct a misconception about diffeomorphisms ...

## General purpose library for multivariate image registration, segmentation & statistical analysis tools

* 170,000+ lines of C++, 6$+$ years of work, 15+ collaborators.

* Generic mathematical methods that are tunable for 
application specific domains:  no-free lunch

* Deep testing on multiple platforms ... osx, linux, windows.

* Several "wins" in public knock-abouts ( [Klein 2009](http://www.ncbi.nlm.nih.gov/pubmed/19195496), [Murphy 2011](http://www.ncbi.nlm.nih.gov/pubmed/21632295), [SATA 2012 and 2013](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3837555/), [BRATS 2013](http://martinos.org/qtim/miccai2013/proc_brats_2013.pdf), others )

```
    An algorithm must use prior knowledge about a problem 
    to do well on that problem 
```

## ANTs: Beyond Registration

![](./figures/antscapabilities.jpg)

[Atropos](http://www.ncbi.nlm.nih.gov/pubmed/?term=atropos+tustison) segmentation, [N4 inhomogeneity correction](http://www.ncbi.nlm.nih.gov/pubmed/?term=N4+tustison), [Eigenanatomy](http://www.ncbi.nlm.nih.gov/pubmed/?term=eigenanatomy+avants), [SCCAN](http://www.ncbi.nlm.nih.gov/pubmed/?term=sparse+canonical+avants), [Prior-constrained PCA](http://www.ncbi.nlm.nih.gov/pubmed/24852460), and [atlas-based label fusion](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC4009425/) and [MALF](http://www.ncbi.nlm.nih.gov/pubmed/21237273) (powerful expert systems for segmentation)

```

## On documentation

![](figures/handsaw.jpg)

simple and nearly self-explanatory

## On documentation

![](figures/Chainsaw.jpg)

documentation is important

## On documentation

![](figures/blindmen.jpg)

... developers can be blind to doc deficiencies

while users are blind to what we provide!

# ANTs Impact

## ANTs Statistics

![](impact/figures/antsvisitors.png)

## ANTs & Neuroscience

We need statistical image analysis <br />at several scales in modern neuroscience

* Macro: *in vivo* structural and functional MRI

* Micro:  high-resolution post-mortem MRI links with in vivo MRI

* Nano: neuron reconstruction ...

* Solutions that are consistent across these scales have the potential
to build multi-scale feature sets or templates and provide new
insights into brain structure and function

* E.g. Parcellation constraints based on histology, tractography, function ...

* Statistical definitions of anatomy/pathology?

* Reinvention of these solutions within each lab ... can we mitigate this?

* Reduce, reuse, recycle ...

## Financial impact

* The **high error rate** in software makes testing for quality assurance critically important.

* Bugs exist **Everywhere**: so don't freak out if we admit we have bugs:  we're just being honest ( and not all devs are )

* Virtually every business in the United States now depends on software for development, production, distribution, and after-sales support of products and services.

* A 2002 NIST study estimated the direct costs to the software supply chain due to failure to identify (**successfully test for**) "bugs"

* We are better at this than most ... **not perfect**

* The estimate of direct costs compiled from industry survey data for the U.S. economy was **\$60 billion per year**

* ... this estimate did not include costs to end users such as lost business (for example, the cost of shutting down the New York Mercantile Exchange in 1998 due to a software failure).

## ANTs: Saving society \$\$\$

starting with neuroscience ...

# Optimal Templates

## "Optimal" templates (?)

* unbiased wrt measurement space

* space is non-linear: reference matters

* encodes prior information: still just averages (usually)

* concept extends across modalities, anatomy, temporality, etc

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.060.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.061.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.062.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.063.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.064.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.065.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.066.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.067.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.068.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.069.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.070.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.071.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.072.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.073.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.074.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.075.jpg)

## Faces, brains, whatever ...

![](templates/figures/WashUJan2011.076.jpg)

see [ANTs site](stnava.github.io/ANTs) and [ANTs Documentation page](stnava.github.io/ANTsDoc)

## Templates from ANTs Colleagues ( not us )

![](templates/figures/templates_ants_colleagues.png)

## Templates unlock modalities

![](figures/ptbp2.png)

# Evaluation

<div align="center"><img width="3500" src="evaluation/figures/evalhistory.png" frameborder="0"></img></div>

# ANTs versus Freesurfer:<br />  Quantifying *life span* brain health

## "Big data" problem from public resources
![](./figures/lifebrains.jpg)

TOT, NKI, IXI, Oasis, ADNI ... several thousand images

## ANTs versus Freesurfer:<br />  Quantifying *life span* brain health

* Freesurfer is the historical standard for measuring cortical thickness

* instead of using surfaces to measure cortical thickness, we use the image space *DiReCTly*

* [see this section of a different talk](http://stnava.github.io/ANTsTalk/#/putting-it-all-together-can-we-quantify-life-span-brain-health-in-individuals-and-in-populations)

* and this "big data" paper: [Large-scale evaluation of ANTs and FreeSurfer cortical thickness measurements](http://www.ncbi.nlm.nih.gov/pubmed/24879923)

* comparison of prediction from automated cortical thickness measurement from 4 public datasets

* $>$ 1200 subjects, age 7 to over 90 years old

* *hint*: ANTs thickness measurements have higher prediction accuracy relative to Freesurfer ( implying we extract more information from the data )

* ANTs methods consistently improve statistical power [eigenanatomy](http://www.ncbi.nlm.nih.gov/pubmed/24687814), [syn](http://www.ncbi.nlm.nih.gov/pubmed/?term=syn+epstein+avants), [itkv4](http://www.ncbi.nlm.nih.gov/pubmed/24817849) ... also, see [Schwarz CG, et al.](http://www.ncbi.nlm.nih.gov/pubmed/24650605) re: TBSS and related work in fMRI [Miller, PNAS](http://www.ncbi.nlm.nih.gov/pubmed/15980148), [Azab, et al in Hippocampus](http://www.ncbi.nlm.nih.gov/pubmed/24167043).

## Cortical thickness estimation

<div align="center"><img height="250" src="./figures/corticalThicknessEstimation.png" frameborder="0"></img></div>

In contrast to FreeSurfer which warps coupled surface meshes to segment the gray matter, *ANTs* diffeomorphically registers the white matter to the combined gray/white matters while simultaneously estimating thickness.

@@

>>>>

@@

## ANTs longitudinal pipeline: TBI

![](longitudinal/figures/long_with_lesions_seg.png)

## ANTs longitudinal pipeline: TBI

![](longitudinal/figures/long_with_lesions_thick.png)

![](longitudinal/figures/babystudy.png)
